Clustering time series based on probability distributions across temporal granularities

### 2021-10-29 main.Rmd
minor changes in validation section

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by ${g10,g11}$, ${g20, g21, g22}$ and ${g30, g31, g32, g33, g34}$ and levels $l_{g_1}=2$, $l_{g_2}=3$ and $l_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We created independent replications $R = \{25, 250, 500\}$ of all data designs to see if our proposed clustering approaches can detect distinct designs in various groups for small, medium and large number of series. A sample size of $T=\{300, 1000, 5000\}$ is used in all designs to test small, medium and large sized series. The methods could preform differently with different jumps between consecutive categories. So a mean difference of $diff = \{1, 2, 5\}$ for corresponding categories are also considered. The performance of the methods can vary with different number of significant granularities. So scenarios with all, few and just one significant granularities are considered. The code for creating these designs and the detailed results can be found in the Supplementary section (link to github repo).

@@

Each category or combination of categories from $g1$, $g2$ and $g3$ are assumed to come from the same distribution, a subset of them from the same distribution, a subset of them from separate distributions, or all from different distributions, resulting in various data designs. As the methods ignore the linear progression of time, there is little value in adding time dependency in the data generating process. It is often reasonable to construct a time series using properties such as trend, seasonality, and auto-correlation. However, when examining distributions across categories of cyclic granularities, these time series features are lost or addressed independently by considering seasonal fluctuations through cyclic granularities. Because the time span during which an entity is observed in order to ascertain its behavior is not very long,
the behavior of the entity will not change drastically and hence the time series can be assumed to remain stationary throughout the observation period. If the observation period is very long (for e.g more than 3 years), property, physical or geographical attributes might change leading to a non-stationary time series. But such a scenario is not considered here and the resulting clusters are assumed to be time invariant in the observation period. The data type is set to be "continuous," and the setup is assumed to be Gaussian. When the distribution of a granularity is "fixed", it means distributions across categories do not vary and are considered to be from N (0,1). The mean of different categories are altered in the "varying" designs, leading to varying distributions across categories.

@@

@@

\noindent Consider the scenario when all three granularities $g1$, $g2$, and $g3$ are responsible for distinguishing the designs. This implies that the patterns across each granularity will change significantly for at least one among the to-be-grouped designs. We consider different distributions across categories (as in Table \ref{tab:tab-dist-design} top) that will lead to different designs (as in Table \ref{tab:tab-dist-design} below). Figure \ref{fig:plot-3gran-new} shows the linear and cyclic representation of the simulated variable under these five designs. As could be seen from the plot, it is impossible to decipher the structural difference in the time series variable just by looking at the linear view. The difference in structure becomes quite clear when we see the distribution across cyclic granularities. Hence, for the consequent scenarios, only graphical displays across cyclic granularities are provided to emphasize the difference in structure.

@@

All the methods were fitted to each data designs and results are reported through confusion matrices. With increasing difference between categories, it gets easier for the methods to correctly distinguish the designs. For $mean_diff=1$, the performances are pretty bad for js-robust methods and wpd method for lower nT. Although, with the kind of residential load datasets, a full year of load is the minimal requirement to capture expected variations in winter and summer profiles, for example. It is likely that $nT$ would be at least $1000$ with half-hourly data, even if data is only available just for a month. The performance is promising except when the number of observations for a customer is really small. For smaller difference between categories, it is expected that method js-nqt would preform better than the other two.
<!-- A confusion table can come alive with linked brushing, so that mismatches and agreements between methods can be explored. -->

>>>>

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $l_{g_1}=2$, $l_{g_2}=3$ and $l_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We created independent replications $R = \{25, 250, 500\}$ of all data designs to see if our proposed clustering approaches can detect distinct designs in various groups for small, medium and large number of series. A sample size of $T=\{300, 1000, 5000\}$ is used in all designs to test small, medium and large sized series. The methods could preform differently with different jumps between consecutive categories. So a mean difference of $diff = \{1, 2, 5\}$ for corresponding categories are also considered. The performance of the methods can vary with different number of significant granularities. So scenarios with all, few and just one significant granularities are considered. The code for creating these designs and the detailed results can be found in the Supplementary section (https://github.com/Sayani07/paper-gracsR).

@@

Each category or combination of categories from $g1$, $g2$ and $g3$ are assumed to come from the same distribution, a subset of them from the same distribution, a subset of them from separate distributions, or all from different distributions, resulting in various data designs. As the methods ignore the linear progression of time, there is little value in adding time dependency in the data generating process. The data type is set to be "continuous," and the setup is assumed to be Gaussian. When the distribution of a granularity is "fixed", it means distributions across categories do not vary and are considered to be from N (0,1). The mean of different categories are altered in the "varying" designs, leading to varying distributions across categories.

@@

<!-- It is often reasonable to construct a time series using properties such as trend, seasonality, and auto-correlation. However, when examining distributions across categories of cyclic granularities, these time series features are lost or addressed independently by considering seasonal fluctuations through cyclic granularities. Because the time span during which an entity is observed in order to ascertain its behavior is not very long, -->
<!-- the behavior of the entity will not change drastically and hence the time series can be assumed to remain stationary throughout the observation period. If the observation period is very long (for e.g more than 3 years), property, physical or geographical attributes might change leading to a non-stationary time series. But such a scenario is not considered here and the resulting clusters are assumed to be time invariant in the observation period.  -->

@@

\noindent Consider the scenario when all three granularities $g1$, $g2$, and $g3$ are responsible for distinguishing the designs. This implies that the patterns across each granularity will change significantly for at least one among the to-be-grouped designs. We consider different distributions across categories (as in Table \ref{tab:tab-dist-design} top) that will lead to different designs (as in Table \ref{tab:tab-dist-design} below). Figure \ref{fig:plot-3gran-new} shows the linear and cyclic representation of the simulated variable under these five designs. As could be seen from the plot, it is impossible to decipher the structural difference in the time series variable just by looking at the linear view. The difference in structure becomes quite clear when we see the distribution across cyclic granularities. Hence, for the consequent scenarios, only graphical displays across cyclic granularities are provided to emphasize the distributional differences in categories.

@@

All the methods were fitted to each data designs and results are reported through cluster statistics or plots. With increasing difference between categories, it gets easier for the methods to correctly distinguish the designs. For ${diff}=1$, the performances are pretty bad for js-robust methods and wpd method for lower nT. Although, with the kind of residential load datasets, a full year of load is the minimal requirement to capture expected variations in winter and summer profiles, for example. It is likely that $nT$ would be at least $1000$ with half-hourly data, even if data is only available just for a month. The performance is promising except when the number of observations for a customer is really small. For smaller difference between categories, it is expected that method js-nqt would preform better than the other two.
<!-- A confusion table can come alive with linked brushing, so that mismatches and agreements between methods can be explored. -->
